---
title: "A07- ggplot"
author: "Suhail M."
date: "June 07 2024"
output: html_notebook
---

# **Recessions and Crime**

### **Guiding Research Question**

The goal of this project is to determine if there is any correlation between a recession and crime rates. In other words, is there more crime during a recession?

In order to do this, I have quarterly GDP data from the St. Louis FED and annual crime statistics from the FBI.

### Data Intake and Overview

#### GDP Data

-   **Source :-** [Real Gross Domestic Product from FRED](https://fred.stlouisfed.org/series/GDPC1)

-   **Origin and Purpose:** Quarterly GDP data sourced from the St. Louis Federal Reserve. This data provides insights into the economic performance of the United States over time.

-   **Variables of Interest:**

    -   `Year`: Year of observation.

    -   `GDP_Rate`: Quarterly GDP growth rate (%).

#### Crime Statistics Data

-   **Source :-** [Crime in the United States from the FBI](https://ucr.fbi.gov/crime-in-the-u.s/2013/crime-in-the-u.s.-2013/tables/1tabledatadecoverviewpdf/table_1_crime_in_the_united_states_by_volume_and_rate_per_100000_inhabitants_1994-2013.xls)

-   **Origin and Purpose:** Annual crime statistics from the FBI's Uniform Crime Reporting (UCR) program. This data captures various crime categories across different years.

-   **Variables of Interest:**

    -   `Year`: Year of observation.

    -   `Violent`, `Murder_and_Manslaughter`, `Rape`, `Robbery`, `Aggravated_Assault`: Total counts of different crime types.

    -   `Violent_Rate`, `Murder_and_Manslaughter_Rate`, etc.: Crime rates per 100,000 population for respective crime types.

### **Examine Each Data Source**

First, I need to load the two data sets into R

```{r}
library(tidyverse)
```

```{r echo=TRUE, message=TRUE}
GDP <- read.csv("C:\\Users\\hp\\Downloads\\MEB\\4007 suhail\\Final-Project-Stat-184\\GDPC1.csv")

# Display structure and summary
str(GDP)
summary(GDP)
head(GDP)
# Check for missing values and convert to numeric
GDP$GDPC1 <- as.numeric(GDP$GDPC1)

# Check for missing values in GDPC1
sum(is.na(GDP$GDPC1))  # Check if there are any NA values

# Plot boxplot if there are no missing values
if (!anyNA(GDP$GDPC1)) {
  boxplot(GDP$GDPC1, main = "Boxplot of Quarterly GDP Growth Rate")
} else {
  cat("There are missing values in GDPC1 column. Please handle them before plotting.")
}
```

```{r}

CRIME <-  read.csv("C:\\Users\\hp\\Downloads\\MEB\\4007 suhail\\Final-Project-Stat-184\\CRIME.csv")

Crime2<-CRIME[-c(1,2,24,25,26,27,28,29,30,31,32,33,34,35,36,37),]  #Delete unnecessary rows from Crime and rename the table

colnames(Crime2)[1] <- "Year"                                      #Renaming the column names because they were wrong
colnames(Crime2)[2] <- "Population"
colnames(Crime2)[3] <- "Violent"
colnames(Crime2)[4] <- "Violent_Rate"
colnames(Crime2)[5] <- "Murder_and_Manslaughter"
colnames(Crime2)[6] <- "Murder_and_Manslaughter_Rate"
colnames(Crime2)[7] <- "Rape"
colnames(Crime2)[8] <- "Rape_Rate"
colnames(Crime2)[9] <- "Robbery"
colnames(Crime2)[10] <- "Robbery_Rate"
colnames(Crime2)[11] <- "Aggravated_Assault"
colnames(Crime2)[12] <- "Aggravated_Assault_Rate"
colnames(Crime2)[13] <- "Property"
colnames(Crime2)[14] <- "Property_Crime"
colnames(Crime2)[15] <- "Burglary"
colnames(Crime2)[16] <- "Burglary_Rate"
colnames(Crime2)[17] <- "Larceny"
colnames(Crime2)[18] <- "Larceny_Rate"
colnames(Crime2)[19] <- "Motor_Vehicle_Theft"
colnames(Crime2)[20] <- "Motor_Vehicle_Theft_Rate"

Crime3<- Crime2[-c(1),]                      #Delete the first row because it had header information instead of data

Crime3$X.19<- NULL                           # Delete the last four columns because they all had NA for data
Crime3$X.20<-NULL
Crime3$X.21 <-NULL
Crime3$X.22<-NULL

Crime3[1] <- lapply(Crime3, as.character)    #convert Year to a character so it can be joinged with GDP Year

CRIME_Final <- Crime3                        #Rename the finished clean table CRIME_Final

head(CRIME_Final)                            #create neat table for CRIME_Final

# Display structure and summary
str(CRIME_Final)
summary(CRIME_Final)

# Convert crime count variables to numeric
CRIME_Final[, c("Violent", "Murder_and_Manslaughter", "Rape", "Robbery", "Aggravated_Assault")] <-
  lapply(CRIME_Final[, c("Violent", "Murder_and_Manslaughter", "Rape", "Robbery", "Aggravated_Assault")],
         function(x) as.numeric(gsub(",", "", x)))

# Convert crime rate variables to numeric
CRIME_Final[, c("Violent_Rate", "Murder_and_Manslaughter_Rate", "Rape_Rate", "Robbery_Rate", "Aggravated_Assault_Rate")] <-
  lapply(CRIME_Final[, c("Violent_Rate", "Murder_and_Manslaughter_Rate", "Rape_Rate", "Robbery_Rate", "Aggravated_Assault_Rate")],
         function(x) as.numeric(x))

# Display structure and summary after conversion
str(CRIME_Final)
summary(CRIME_Final)

# Check for outliers in crime rates
boxplot(CRIME_Final$Violent_Rate, main = "Boxplot of Violent Crime Rate")
```

### **Graphics**

Time Series Plot of GDP Growth Rate

```{r}
# Convert DATE to Date format
GDP$DATE <- as.Date(GDP$DATE)

# Load ggplot2 package
library(ggplot2)

# Create the time series plot
ggplot(GDP, aes(x = DATE, y = GDPC1)) +
  geom_line() +
  labs(title = "US Quarterly GDP Growth Rate",
       x = "Date",
       y = "Growth Rate (%)")
```

Violent Crime Rate Over Time

```{r}
CRIME_Final$Year <- as.numeric(CRIME_Final$Year)
CRIME_Final$Violent_Rate <- as.numeric(gsub(",", "", CRIME_Final$Violent_Rate))  # Remove commas and convert to numeric

# Create the line plot
ggplot(CRIME_Final, aes(x = Year, y = Violent_Rate)) +
  geom_line() +
  labs(title = "US Violent Crime Rate",
       x = "Year",
       y = "Rate per 100,000 Population")
```

In order to find which quarters were in a recession, I will calculate the percentage change for GDP. Then I can filter out all of the positive values to get only recession entries.

```{r}
#calculate pecentage change of GDP and only display the values for a negative value to show recession years. 

GDP_Recession <- GDP%>%                                    #name the table GDP_Recessions and use the GDP data
  mutate(pChange=(GDPC1-lag(GDPC1))/lag(GDPC1)*100)%>% #creates the percentage change variable
  filter(pChange<0) %>%                                #Only display the negative percentage change values
  na.omit()                                            #omit any entries that say NA

head(GDP_Recession)
```

Now I will finalize my GDP data. GDP is reported quarterly, but I need to make it annual so that it will match with the Crime table. To make it annual data, I am averaging the percentage change values for each year.

```{r echo=TRUE}
#Crime data is only yearly. so need to get the GDP_Recession data into yearly. To do this, find the yearly average rate for GDP_Recession. Probably better to tell an actual recession because it'll give a closer estimate of consecutive periods and official recessions. 

library(stringi)                                         #Required for stri_extract_last_regex

GDP_Recession2 <- GDP_Recession %>%                      #Extracts the year (last four digits) from the date
  mutate(Year=(stri_extract_last_regex(DATE, "\\d{4}")))

GDP_Final<-
group_by(GDP_Recession2,Year ) %>%                       #Finds the average annual percent change in GDP
  summarize(GDP_Rate= mean(pChange))


#create neat table for GDP_FinalNext, I need to take some steps to clean my Crime data-set so that R will be able to use it
head(GDP_Final) 
```

Now I have two tables that are ready to be joined. I will use the year to match them so that each case is a year with crime statistics and a GDP percentage change value

```{r}
#Need to Join the two tables so each year has a percentage change and crime statistics

# Convert GDP_Final$Year to numeric
GDP_Final$Year <- as.numeric(GDP_Final$Year)

# Now perform the full join
Total <- CRIME_Final %>%                #Join CRIME_Final and GDP_Final by year
  full_join(GDP_Final, by = "Year") %>%
  na.omit()

head(Total)             #create neat table for Total
```

The tables are joined and now I can plot all the data to see if there is any correlation.\
Address Research Question

### **Based on the initial EDA:**

-   The graphics will help us to visualize any potential patterns or trends in GDP growth rates and crime rates.

-   Observations from these plots will guide us further analysis into the correlation between recessions and crime rates.
